Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data

### 2025-01-16 paper/paper.tex
Added TODOs
Added 3 models results

\subsection{Fairness in AI}

@@

\subsection{Fairness Auditing}

@@

\subsection{Challenges in Auditing}

@@

\subsection{Differentially Privacy}

@@

\subsection{Synthetic Data}

Synthetic data are fake data that is generated from real data. Synthetic data is artificially generated to replicate the statistical properties of real data while protecting privacy. The principle of this technology is to preserve some statistical properties of the real data while protecting the privacy of the individuals in the real data. Unlike anonymized data, synthetic data breaks the link to the original data, reducing the risk of re-identification.

@@

\subsection{Contribution}

@@

\subsection{Structure}

@@

@@

Next, several AI models are extracted from various authors from Kaggle. They are finetuned to perform well on the original dataset. For one, a random forest model is finetuned by searching hyperparameters settings\cite{Ipbyrne2023}. Another random forest model is finetuned with over-sampling methods\cite{Jawat2024}. Also, a logistic regression model is finetuned by performing principal component analysis\cite{Prashant1112023}.

The AI models and both the original dataset and synthetic datasets are then fed to the fairness checker. The synthetic datasets were generated 100 times and the fairness scores averaged across the 100 runs. Sensitive attributes are identified based on manual examination with common sense or by referring to \cite{pessach2022review}. Then, applicable fairness measures are computed using the checker for both the original and the synthetic.

@@

The Adult\cite{adult_2,Kaggle_Adult_Census_Income} dataset comes from the 1994 census in the United States and contains about 30000 individuals. The dataset has 14 attributes, such as sex, age, and education. The task is to predict whether an individual earns more than \$50,000 a year. We set the sensitive attribute as sex and the protected group as female. The results are shown in Table~\ref{tab:adult_score}.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Adult dataset.
    % With heuristic, average difference is $0.0787$, and fit time is 924m15s.
    Average difference is $0.0775$, and fit time is 17m1s.
}
% \label{tab:adult_score}
% \begin{tabular}{ccccccc}
% \toprule
% \textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{difference} & \textbf{Synthetic(No Heuristic)} & \textbf{difference(No Heuristic)} \\
% \midrule
% Demographic Parity  & 0.1933 & 0.1079 & 0.0853 & 0.0114 & 0.1818 \\
% Overall Accuracy Equality   & 0.0250 & 0.1399 & 0.1149 & 0.0169 & 0.0080 \\
% Equalized Odds (False Positive)    & 0.0175 & 0.0870 & 0.0695 & 0.0060 & 0.0115 \\
% Equalized Odds (True Positive)    & 0.0114 & 0.0622 & 0.0508 & 0.0303 & 0.0189 \\
% CU Accuracy Equality (True Positive) & 0.1941 & 0.1466 & 0.0474 & 0.0315 & 0.1626 \\
% CU Accuracy Equality (True Negative) & 0.1102 & 0.1450 & 0.0348 & 0.0299 & 0.0803 \\
% \bottomrule
% \end{tabular}
% \end{table}

\label{tab:adult_score}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.1933 & 0.0114 & 0.1818 \\
Overall Accuracy Equality   & 0.0250 & 0.0169 & 0.0080 \\
Equalized Odds (False Positive)    & 0.0175 & 0.0060 & 0.0115 \\
Equalized Odds (True Positive)    & 0.0114 & 0.0303 & 0.0189 \\
Conditional Use Accuracy Equality (True Positive) & 0.1941 & 0.0315 & 0.1626 \\
Conditional Use Accuracy Equality (True Negative) & 0.1102 & 0.0299 & 0.0803 \\
\bottomrule
\end{tabular}
\end{table}

@@

The COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset} dataset comes from an investigative report by ProPublica of the COMPAS criminal recidivism assessment system and contains about 7000 individuals. The dataset has 10 applicable attributes, such sex, age, and priors count. The task is to predict whether an individual will re-offend. We set the sensitive attribute as sex and the protected group as female. The results are shown in Table~\ref{tab:compas_score}.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the COMPAS dataset.
    % With heuristic, average difference is $0.0750$, and fit time is 27m33s.
    Average difference is $0.0727$, and fit time is 33s.
}
\label{tab:compas_score}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.1310 & 0.0809 & 0.0501 \\
Overall Accuracy Equality   & 0.0079 & 0.0146 & 0.0067 \\
Equalized Odds (False Positive)    & 0.0249 & 0.0802 & 0.0553 \\
Equalized Odds (True Positive)    & 0.0177 & 0.0825 & 0.0648 \\
Conditional Use Accuracy Equality (True Positive) & 0.1709 & 0.0551 & 0.1157 \\
Conditional Use Accuracy Equality (True Negative) & 0.1695 & 0.0256 & 0.1439 \\
\bottomrule
\end{tabular}
\end{table}

@@

The Diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction} dataset comes from the hospital readmission data published in the 1994 AI in Medicine journal and contains about 100000 individuals. The dataset has 8 attributes, such as gender, age, and smoking history. The task is to predict whether an individual will be readmitted. We set the sensitive attribute as gender and the protected group as female. The results are shown in Table~\ref{tab:diabetes_score}.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Diabetes dataset.
    % With heuristic, average difference is $0.0067$, and fit time is 1m44s.
    Average difference is $0.0078$, and fit time is 1m24s.
}
\label{tab:diabetes_score}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.0135 & 0.0015 & 0.0119 \\
Overall Accuracy Equality   & 0.0077 & 0.0011 & 0.0065 \\
Equalized Odds (False Positive)    & 0.0000 & 0.0010 & 0.0010 \\
Equalized Odds (True Positive)    & 0.0086 & 0.0094 & 0.0008 \\
Conditional Use Accuracy Equality (True Positive) & 0.0133 & 0.0059 & 0.0074 \\
Conditional Use Accuracy Equality (True Negative) & 0.0216 & 0.0019 & 0.0197 \\
\bottomrule
\end{tabular}
\end{table}

@@

Future work could focus on applying this framework to larger, more complex datasets, which would help validate the framework's scalability. Furthermore, other privacy-preserving techniques could be explored to complement synthetic data and expand the scope of AI fairness auditing while maintaining privacy. Finally, translating these technical advancements into real-world applications requires policy, legal, and industry collaborations. Research on the legal, ethical, and societal implications of differential privacy and synthetic data will be crucial for broad adoption.

>>>>

% \subsection{Fairness in AI}

@@

% \subsection{Fairness Auditing}

@@

% \subsection{Challenges in Auditing}

@@

% \subsection{Differentially Privacy}

@@

% \subsection{Synthetic Data}

One way of auditing while preserving privacy is to use synthetic data. Synthetic data are fake data that is generated from real data. Synthetic data is artificially generated to replicate the statistical properties of real data while protecting privacy. The principle of this technology is to preserve some statistical properties of the real data while protecting the privacy of the individuals in the real data. Unlike anonymized data, synthetic data breaks the link to the original data, reducing the risk of re-identification.

@@

% \subsection{Contribution}

@@

% \subsection{Structure}

@@

% TODO: add to evaluation
% TODO: add to conclusion

@@

Next, several AI models are extracted from various authors from Kaggle. They are finetuned to perform well on the original dataset. For the Adult dataset, we selected three models to experiment: a random forest model is finetuned by hyperparameter search\cite{Ipbyrne2023}; a logistic regression model is finetuned by performing principal component analysis\cite{Prashant1112023}; and a KNN model is manually finetuned\cite{Kothari2020}. For the COMPAS dataset, we selected a random forest model finetuned by hyperparameter search\cite{Sky2024}. For the Diabetes dataset, we selected a random forest model finetuned by over-sampling and under-sampling techniques\cite{Jawat2024}.

The AI models and both the original dataset and synthetic datasets are then fed to the fairness checker. The synthetic datasets were generated 100 times for each model and the fairness scores averaged across the 100 runs. Sensitive attributes are identified based on manual examination with common sense or by referring to \cite{pessach2022review}. Then, applicable fairness measures are computed using the checker for both the original and the synthetic.

@@

The Adult\cite{adult_2,Kaggle_Adult_Census_Income} dataset comes from the 1994 census in the United States and contains about 30000 individuals. The dataset has 14 attributes, such as sex, age, and education. The task is to predict whether an individual earns more than \$50,000 a year. We set the sensitive attribute as sex and the protected group as female. We tested three models: a random forest model, a logistic regression model, and a KNN model. The results are shown in Table~\ref{tab:adult_score_rf}, \ref{tab:adult_score_lr}, and \ref{tab:adult_score_knn}. The synthetic data generator fit time is 17 minutes and 1 second.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Adult dataset on a random forest model.
    % With heuristic, average difference is $0.0787$, and fit time is 924m15s.
    Average difference is $0.0775$.
}
\label{tab:adult_score_rf}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.1933 & 0.0114 & 0.1818 \\
Overall Accuracy Equality   & 0.0250 & 0.0169 & 0.0080 \\
Equalized Odds (False Positive)    & 0.0175 & 0.0060 & 0.0115 \\
Equalized Odds (True Positive)    & 0.0114 & 0.0303 & 0.0189 \\
Conditional Use Accuracy Equality (True Positive) & 0.1941 & 0.0315 & 0.1626 \\
Conditional Use Accuracy Equality (True Negative) & 0.1102 & 0.0299 & 0.0803 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Adult dataset on a logistic regression model.
    Average difference is $0.0829$.
}
\label{tab:adult_score_lr}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.0139 & 0.0047 & 0.0092 \\
Overall Accuracy Equality   & 0.0992 & 0.0149 & 0.0843 \\
Equalized Odds (False Positive)    & 0.0017 & 0.0066 & 0.0049 \\
Equalized Odds (True Positive)    & 0.0080 & 0.0072 & 0.0008 \\
Conditional Use Accuracy Equality (True Positive) & 0.3459 & 0.0305 & 0.3154 \\
Conditional Use Accuracy Equality (True Negative) & 0.1120 & 0.0291 & 0.0829 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Adult dataset on a KNN model.
    Average difference is $0.0945$.
}
\label{tab:adult_score_knn}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity & 0.1721 & 0.0046 & 0.1675 \\
Overall Accuracy Equality &  0.0473 & 0.0128 & 0.0344 \\
Equalized Odds (False Positive) &   0.0571 & 0.0072 & 0.0499 \\
Equalized Odds (True Positive) &   0.1667 & 0.0110 & 0.1557 \\
Conditional Use Accuracy Equality (True Positive) & 0.1009 & 0.0288 & 0.0720 \\
Conditional Use Accuracy Equality (True Negative) & 0.1193 & 0.0319 & 0.0874 \\
\bottomrule
\end{tabular}
\end{table}

@@

The COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset} dataset comes from an investigative report by ProPublica of the COMPAS criminal recidivism assessment system and contains about 7000 individuals. The dataset has 10 applicable attributes, such sex, age, and priors count. The task is to predict whether an individual will re-offend. We set the sensitive attribute as sex and the protected group as female. The results are shown in Table~\ref{tab:compas_score}. The synthetic data generator fit time is 33 seconds.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the COMPAS dataset on a random forest model.
    % With heuristic, average difference is $0.0750$, and fit time is 27m33s.
    Average difference is $0.0727$.
}
\label{tab:compas_score}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.1310 & 0.0809 & 0.0501 \\
Overall Accuracy Equality   & 0.0079 & 0.0146 & 0.0067 \\
Equalized Odds (False Positive)    & 0.0249 & 0.0802 & 0.0553 \\
Equalized Odds (True Positive)    & 0.0177 & 0.0825 & 0.0648 \\
Conditional Use Accuracy Equality (True Positive) & 0.1709 & 0.0551 & 0.1157 \\
Conditional Use Accuracy Equality (True Negative) & 0.1695 & 0.0256 & 0.1439 \\
\bottomrule
\end{tabular}
\end{table}

@@

The Diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction} dataset comes from the hospital readmission data published in the 1994 AI in Medicine journal and contains about 100000 individuals. The dataset has 8 attributes, such as gender, age, and smoking history. The task is to predict whether an individual will be readmitted. We set the sensitive attribute as gender and the protected group as female. The results are shown in Table~\ref{tab:diabetes_score}. The synthetic data generator fit time is 1 minute and 24 seconds.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Diabetes dataset on a random forest model.
    % With heuristic, average difference is $0.0067$, and fit time is 1m44s.
    Average difference is $0.0078$.
}
\label{tab:diabetes_score}
\begin{tabular}{lccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} \\
\midrule
Demographic Parity  & 0.0135 & 0.0015 & 0.0119 \\
Overall Accuracy Equality   & 0.0077 & 0.0011 & 0.0065 \\
Equalized Odds (False Positive)    & 0.0000 & 0.0010 & 0.0010 \\
Equalized Odds (True Positive)    & 0.0086 & 0.0094 & 0.0008 \\
Conditional Use Accuracy Equality (True Positive) & 0.0133 & 0.0059 & 0.0074 \\
Conditional Use Accuracy Equality (True Negative) & 0.0216 & 0.0019 & 0.0197 \\
\bottomrule
\end{tabular}
\end{table}

@@

Future work could focus on applying this framework to larger, more complex datasets, which would help validate the framework's scalability. Furthermore, other privacy-preserving techniques could be explored to complement synthetic data and expand the scope of AI fairness auditing while maintaining privacy. Other means of auditing for certain fairness metrics, such as Conditional Use Accuracy Equality, that are not well suited for synthetic data can be explored. Finally, translating these technical advancements into real-world applications requires policy, legal, and industry collaborations. Research on the legal, ethical, and societal implications of differential privacy and synthetic data will be crucial for broad adoption.
